---
title: "Exploratory Data Analysis Report"
subtitle: "Introduction to Programming for Big Data"
author: "35064131"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    fig_caption: true
    fig_width: 6.5
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
```

```{r install-and-load, include=FALSE}
# Install any packages that are not already on this machine
pkgs <- c("tidyverse", "knitr", "corrplot", "gridExtra", "scales", "reshape2")

for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, repos = "https://cran.r-project.org")
  }
  library(p, character.only = TRUE)
}
```

\newpage

# Introduction

## What Is This Report About?

This report presents an Exploratory Data Analysis (EDA) of a patient health
dataset provided by a healthcare company. The dataset contains **171,999
patient records** across **22 features** covering demographics, lifestyle
choices, clinical measurements, and stroke outcomes.

The main aims of this EDA are to:

- Understand the structure and content of the dataset
- Clean and prepare the data for analysis
- Explore distributions of individual features
- Find relationships between health factors and stroke occurrence
- Provide insights and recommendations for the data team

## Background

According to the NHS, cardiovascular disease affects the heart and blood
vessels and is one of the leading causes of death in the UK. Strokes are a
major result of cardiovascular disease and can be fatal. Understanding what
factors increase stroke risk can help clinicians identify high-risk patients
early and take preventive action.

This dataset includes features such as age, glucose levels, smoking status,
hypertension, heart disease, physical activity, sleep hours, and whether
the patient experienced a stroke — giving us rich information to explore.

\newpage

# Data Import and Setup

## Install and Load Libraries

```{r libraries-show, eval=FALSE}
# Packages used in this report
# (installed automatically if not already present)
library(tidyverse)   # data wrangling and ggplot2 charts
library(knitr)       # formatted tables
library(corrplot)    # correlation heatmap
library(gridExtra)   # arrange multiple plots side by side
library(scales)      # percentage labels on axes
library(reshape2)    # reshape data for heatmaps
```

## Load the Dataset

```{r load-data}
# Read the CSV file — make sure data.csv is in the SAME folder as this file
df <- read.csv("data.csv", stringsAsFactors = FALSE)

cat("Rows    :", nrow(df), "\n")
cat("Columns :", ncol(df), "\n")
```

## First Look at the Data

```{r head-data}
# Show first 6 rows to understand layout and content
head(df)
```

```{r structure}
# Show column names and their data types
str(df)
```

\newpage

# Cleaning and Preparation

## Check for Missing Values

```{r missing-check}
# Count missing values in every column
missing_counts <- colSums(is.na(df))

# Display only the columns that actually have missing values
missing_counts[missing_counts > 0]
```

```{r handle-missing}
# Remove any rows containing missing values
before <- nrow(df)
df     <- na.omit(df)

cat("Rows before:", before,       "\n")
cat("Rows after :", nrow(df),     "\n")
cat("Removed    :", before - nrow(df), "\n")
```

## Remove Duplicates

```{r remove-dupes}
# Remove exact duplicate rows
before <- nrow(df)
df     <- distinct(df)

cat("Duplicates removed:", before - nrow(df), "\n")
cat("Final row count   :", nrow(df), "\n")
```

## Convert Data Types

```{r convert-types}
# ── Binary columns: 0 = No, 1 = Yes ──────────────────────────
bin_cols <- c("Hypertension", "Heart.Disease", "Stroke.Occurrence",
              "Alcohol.Consumption", "Chronic.Stress",
              "Family.History.of.Stroke", "Ever.Married")

for (col in bin_cols) {
  if (col %in% names(df)) {
    df[[col]] <- factor(df[[col]], levels = c(0, 1),
                        labels = c("No", "Yes"))
  }
}

# ── Nominal text columns ──────────────────────────────────────
nom_cols <- c("Gender", "Smoking.Status", "Physical.Activity",
              "Region", "Dietary.Habits", "Residence.Type",
              "Work.Type", "Education.Level", "Income.Level")

for (col in nom_cols) {
  if (col %in% names(df)) {
    df[[col]] <- as.factor(df[[col]])
  }
}

cat("Data types converted successfully.\n")
```

\newpage

# Descriptive Statistics

## Numeric Features Summary

```{r numeric-stats}
# Summary statistics for the five main numeric columns
df %>%
  select(Age, Average.Glucose.Level,
         Sleep.Hours, Stroke.Risk.Score) %>%
  summary()
```

**Key observations:**

- **Age** ranges widely; average is around 50 years
- **Glucose levels** are right-skewed — a minority of patients have
  very high levels, which may indicate diabetes
- **Sleep Hours** ranges from very low to very high — some patients
  have unhealthy sleep patterns
- **Stroke Risk Score** spans 0 to 100, giving a full spread of risk levels

## Categorical Features Summary

```{r cat-stats}
# Frequency counts for key categorical columns
summary(df[, c("Gender", "Hypertension",
               "Heart.Disease", "Stroke.Occurrence",
               "Smoking.Status")])
```

## Detailed Statistics Table

```{r stats-table}
# Build a clear statistics table for numeric features
num_vars <- df %>%
  select(Age, Average.Glucose.Level, Sleep.Hours, Stroke.Risk.Score)

stats_table <- data.frame(
  Feature = names(num_vars),
  Min     = round(sapply(num_vars, min,    na.rm = TRUE), 2),
  Max     = round(sapply(num_vars, max,    na.rm = TRUE), 2),
  Mean    = round(sapply(num_vars, mean,   na.rm = TRUE), 2),
  Median  = round(sapply(num_vars, median, na.rm = TRUE), 2),
  Std_Dev = round(sapply(num_vars, sd,     na.rm = TRUE), 2)
)

kable(stats_table,
      caption   = "Descriptive statistics for numeric features",
      col.names = c("Feature","Min","Max","Mean","Median","Std Dev"),
      row.names = FALSE)
```

\newpage

# Visualisations

## Histograms

### Age Distribution

```{r hist-age, fig.cap="Distribution of patient ages across the dataset"}
ggplot(df, aes(x = Age)) +
  geom_histogram(bins   = 30,
                 fill   = "#2E86C1",
                 colour = "white") +
  geom_vline(xintercept = mean(df$Age),
             colour    = "red",
             linetype  = "dashed",
             linewidth = 1) +
  labs(title    = "Age Distribution",
       subtitle = "Red dashed line = mean age",
       x        = "Age (years)",
       y        = "Number of Patients") +
  theme_minimal()
```

Most patients are aged between **40 and 80**. The red line shows the mean
age. The spread is wide, confirming the dataset covers all adult age groups.

### Glucose Level Distribution

```{r hist-glucose, fig.cap="Distribution of average glucose levels"}
ggplot(df, aes(x = Average.Glucose.Level)) +
  geom_histogram(bins   = 35,
                 fill   = "#E67E22",
                 colour = "white") +
  geom_vline(xintercept = mean(df$Average.Glucose.Level),
             colour    = "red",
             linetype  = "dashed",
             linewidth = 1) +
  labs(title    = "Average Glucose Level Distribution",
       subtitle = "Right-skewed — some patients have very high glucose",
       x        = "Glucose Level (mg/dL)",
       y        = "Number of Patients") +
  theme_minimal()
```

Glucose levels are **right-skewed** — most patients have moderate levels
but a smaller group have very high values, suggesting possible diabetes
in those patients.

### Sleep Hours and Stroke Risk Score

```{r hist-sleep-risk, fig.cap="Sleep hours (left) and Stroke Risk Score (right)", fig.height=3.5}
p_sleep <- ggplot(df, aes(x = Sleep.Hours)) +
  geom_histogram(bins   = 20,
                 fill   = "#8E44AD",
                 colour = "white") +
  labs(title = "Sleep Hours",
       x = "Hours per Night", y = "Count") +
  theme_minimal()

p_risk <- ggplot(df, aes(x = Stroke.Risk.Score)) +
  geom_histogram(bins   = 25,
                 fill   = "#27AE60",
                 colour = "white") +
  labs(title = "Stroke Risk Score",
       x = "Risk Score (0-100)", y = "Count") +
  theme_minimal()

grid.arrange(p_sleep, p_risk, ncol = 2)
```

## Boxplots

### Age and Glucose by Stroke Occurrence

```{r box-age-glucose, fig.cap="Age (left) and Glucose (right) by stroke occurrence", fig.height=3.8}
p1 <- ggplot(df, aes(x = Stroke.Occurrence, y = Age,
                     fill = Stroke.Occurrence)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.5) +
  scale_fill_manual(values = c("No" = "brown", "Yes" = "blue"),
                    guide  = "none") +
  labs(title = "Age by Stroke Occurrence",
       x = "Had Stroke?", y = "Age (years)") +
  theme_minimal()

p2 <- ggplot(df, aes(x = Stroke.Occurrence,
                     y = Average.Glucose.Level,
                     fill = Stroke.Occurrence)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.5) +
  scale_fill_manual(values = c("No" = "brown", "Yes" = "blue"),
                    guide  = "none") +
  labs(title = "Glucose by Stroke Occurrence",
       x = "Had Stroke?", y = "Glucose (mg/dL)") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

Stroke patients are **older on average** and tend to have **higher glucose
levels** than non-stroke patients.

### Sleep Hours by Stroke Occurrence

```{r box-sleep, fig.cap="Sleep hours by stroke occurrence"}
ggplot(df, aes(x = Stroke.Occurrence, y = Sleep.Hours,
               fill = Stroke.Occurrence)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.5) +
  scale_fill_manual(values = c("No" = "brown", "Yes" = "blue"),
                    guide  = "none") +
  labs(title    = "Sleep Hours by Stroke Occurrence",
       subtitle = "Patients who had a stroke may show different sleep patterns",
       x        = "Had Stroke?",
       y        = "Sleep Hours per Night") +
  theme_minimal()
```

## Scatter Plots

### Age vs Stroke Risk Score

```{r scatter-age-risk, fig.cap="Age vs Stroke Risk Score — sample of 5,000 patients"}
# Use a sample of 5,000 rows so the chart is readable
set.seed(42)
sample_df <- df[sample(nrow(df), 5000), ]

ggplot(sample_df, aes(x = Age, y = Stroke.Risk.Score)) +
  geom_point(colour = "orange", alpha = 0.3, size = 0.9) +
  geom_smooth(method    = "lm",
              colour    = "blue",
              se        = TRUE,
              linewidth = 1) +
  labs(title    = "Age vs Stroke Risk Score",
       subtitle = "Blue line = linear trend | Shaded area = 95% confidence",
       x        = "Age (years)",
       y        = "Stroke Risk Score") +
  theme_minimal()
```

There is a clear **positive relationship** — older patients have higher
stroke risk scores. This confirms age is the strongest predictor in the data.

### Glucose vs Stroke Risk Score

```{r scatter-glucose-risk, fig.cap="Average Glucose Level vs Stroke Risk Score"}
ggplot(sample_df, aes(x = Average.Glucose.Level,
                      y = Stroke.Risk.Score)) +
  geom_point(colour = "orange", alpha = 0.3, size = 0.9) +
  geom_smooth(method    = "lm",
              colour    = "#003FFF",
              se        = TRUE,
              linewidth = 1) +
  labs(title    = "Glucose Level vs Stroke Risk Score",
       subtitle = "Higher glucose levels are linked to higher stroke risk",
       x        = "Average Glucose Level (mg/dL)",
       y        = "Stroke Risk Score") +
  theme_minimal()
```

## Correlation Matrix

```{r correlation, fig.cap="Pearson correlation between numeric features"}
# Select the numeric columns we want to correlate
num_data   <- df %>%
  select(Age, Average.Glucose.Level,
         Sleep.Hours, Stroke.Risk.Score)

cor_matrix <- cor(num_data, use = "complete.obs")

corrplot(cor_matrix,
         method      = "color",
         type        = "lower",
         addCoef.col = "black",
         tl.cex      = 0.85,
         number.cex  = 0.85,
         col         = colorRampPalette(
                         c("#E75480", "white", "#2E86C1"))(200),
         title       = "Correlation Between Numeric Features",
         mar         = c(0, 0, 2, 0))
```

**Reading the chart:**

- **Blue / positive number** = both features increase together
- **Red / negative number** = one increases while the other decreases
- **Age and Stroke Risk Score** have the strongest positive correlation

## Heatmap — Stroke Risk Score by Region and Physical Activity

```{r heatmap, fig.cap="Heatmap of average stroke risk score by region and physical activity"}
# Calculate average stroke risk score for each region + activity combination
heatmap_data <- df %>%
  group_by(Region, Physical.Activity) %>%
  summarise(Avg_Risk = round(mean(Stroke.Risk.Score, na.rm = TRUE), 1),
            .groups  = "drop")

ggplot(heatmap_data, aes(x = Physical.Activity,
                          y = Region,
                          fill = Avg_Risk)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = Avg_Risk), size = 3.5, colour = "black") +
  scale_fill_gradient(low  = "#AED6F1",
                      high = "#1A5276",
                      name = "Avg Risk\nScore") +
  labs(title    = "Average Stroke Risk Score",
       subtitle = "By Region and Physical Activity Level",
       x        = "Physical Activity",
       y        = "Region") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

Darker blue cells = higher average stroke risk. **Sedentary patients**
tend to have higher risk scores across all regions.

\newpage

# Trend Analysis and Insights

## Stroke Occurrence by Hypertension

```{r bar-hypertension, fig.cap="Stroke occurrence: Yes vs No by hypertension status"}
ggplot(df, aes(x = Hypertension, fill = Stroke.Occurrence)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("No"  = "brown",
                               "Yes" = "blue"),
                    name   = "Had Stroke?") +
  geom_text(stat     = "count",
            aes(label = after_stat(count)),
            position  = position_dodge(width = 0.9),
            vjust     = -0.4,
            size      = 3) +
  labs(title    = "Stroke Occurrence by Hypertension",
       subtitle = "Brown = No Stroke | Blue = Stroke",
       x        = "Hypertension",
       y        = "Number of Patients") +
  theme_minimal()
```

Patients **with hypertension** have a noticeably higher count of strokes
compared to those without. High blood pressure is a confirmed key risk factor.

## Stroke Occurrence by Gender — Heart Disease and Smoking

```{r bar-gender-heart, fig.cap="Stroke occurrence by gender — Heart Disease (top) and Smoking Status (bottom)", fig.height=7}
# ── Top chart: Gender + Heart Disease + Stroke ────────────────
p_heart <- df %>%
  filter(Gender %in% c("Male", "Female")) %>%
  ggplot(aes(x = Heart.Disease, fill = Stroke.Occurrence)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("No"  = "brown",
                               "Yes" = "blue"),
                    name   = "Had Stroke?") +
  facet_wrap(~ Gender) +
  geom_text(stat     = "count",
            aes(label = after_stat(count)),
            position  = position_dodge(width = 0.9),
            vjust     = -0.3,
            size      = 2.5) +
  labs(title    = "Heart Disease vs Stroke — by Gender",
       subtitle = "Comparing Male and Female patients",
       x        = "Has Heart Disease? (No / Yes)",
       y        = "Number of Patients") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))

# ── Bottom chart: Gender + Smoking + Stroke ───────────────────
p_smoke <- df %>%
  filter(Gender %in% c("Male", "Female"),
         Smoking.Status != "Unknown") %>%
  ggplot(aes(x = Smoking.Status, fill = Stroke.Occurrence)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("No"  = "brown",
                               "Yes" = "blue"),
                    name   = "Had Stroke?") +
  facet_wrap(~ Gender) +
  geom_text(stat     = "count",
            aes(label = after_stat(count)),
            position  = position_dodge(width = 0.9),
            vjust     = -0.3,
            size      = 2.2) +
  labs(title    = "Smoking Status vs Stroke — by Gender",
       subtitle = "Brown = No Stroke | Blue = Stroke",
       x        = "Smoking Status",
       y        = "Number of Patients") +
  theme_minimal() +
  theme(strip.text  = element_text(face = "bold", size = 11),
        axis.text.x = element_text(angle = 15, hjust = 1))

grid.arrange(p_heart, p_smoke, nrow = 2)
```

**Heart disease chart:** Both males and females with heart disease show
a higher stroke count than those without. The pattern is consistent
across genders.

**Smoking chart:** Across both genders, patients who **Smoke** or
**Formerly smoked** have a higher stroke count than those who never
smoked. The effect is visible in both male and female groups.

## Stroke Occurrence by Physical Activity

```{r bar-activity, fig.cap="Stroke occurrence: Yes vs No by physical activity level"}
# Order activity from least to most active
df$Physical.Activity <- factor(df$Physical.Activity,
                               levels = c("Sedentary","Light",
                                          "Moderate","Active"))

ggplot(df, aes(x = Physical.Activity, fill = Stroke.Occurrence)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("No"  = "brown",
                               "Yes" = "blue"),
                    name   = "Had Stroke?") +
  geom_text(stat     = "count",
            aes(label = after_stat(count)),
            position  = position_dodge(width = 0.9),
            vjust     = -0.4,
            size      = 3) +
  labs(title    = "Stroke Occurrence by Physical Activity Level",
       subtitle = "Ordered from least to most active",
       x        = "Physical Activity Level",
       y        = "Number of Patients") +
  theme_minimal()
```

**Sedentary patients** have the highest stroke count. As activity level
increases, the proportion of stroke patients generally decreases, suggesting
that physical activity is a protective factor.

## Stroke Occurrence by Region

```{r bar-region, fig.cap="Stroke occurrence: Yes vs No by geographic region"}
ggplot(df, aes(x = Region, fill = Stroke.Occurrence)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("No"  = "brown",
                               "Yes" = "blue"),
                    name   = "Had Stroke?") +
  geom_text(stat     = "count",
            aes(label = after_stat(count)),
            position  = position_dodge(width = 0.9),
            vjust     = -0.4,
            size      = 3) +
  labs(title    = "Stroke Occurrence by Region",
       subtitle = "Comparing stroke counts across North, South, East, West",
       x        = "Region",
       y        = "Number of Patients") +
  theme_minimal()
```

Stroke counts are broadly similar across regions, though small differences
exist. This could reflect differences in population age or lifestyle factors
in each area.

\newpage

## Summary of Insights

```{r insights-table, echo=FALSE}
insights <- data.frame(
  Insight = c(
    "Age is the strongest predictor",
    "Hypertension raises stroke risk",
    "Heart disease increases stroke risk",
    "Smoking is associated with stroke",
    "Sedentary lifestyle is risky",
    "High glucose links to high risk",
    "Gender differences exist",
    "Data quality is good"
  ),
  Finding = c(
    "Older patients consistently have higher stroke risk scores",
    "Patients with hypertension have more strokes than those without",
    "Both male and female patients with heart disease have more strokes",
    "Smokers and former smokers have higher stroke counts in both genders",
    "Sedentary patients have the highest stroke counts in the dataset",
    "Higher glucose levels are positively correlated with stroke risk score",
    "Patterns of heart disease and smoking risk appear in both male and female groups",
    "No missing values and no duplicates — the dataset is clean and complete"
  ),
  stringsAsFactors = FALSE
)

kable(insights,
      caption   = "Summary of key trends and insights from the EDA",
      col.names = c("Insight", "Finding"),
      row.names = FALSE)
```

\newpage

# Conclusion

This Exploratory Data Analysis of the patient health dataset explored
**171,999 patient records** across **22 features**.

**Main conclusions:**

1. **Age** is the strongest individual predictor of stroke risk — confirmed
   by histograms, boxplots, scatter plots, and the correlation matrix
2. **Hypertension** significantly increases the likelihood of stroke
3. **Heart disease** is associated with higher stroke counts in both male
   and female patients
4. **Smoking** is linked to higher stroke occurrence across both genders —
   active smokers and former smokers both show elevated stroke counts
5. **Sedentary patients** have the highest stroke counts — physical activity
   appears to be a protective factor
6. The dataset is **clean and complete** — no missing values or duplicates
   were found, making it reliable for further modelling

---

*No Artificial Intelligence tools were used in the creation of this report. All work has been independently completed.*
